Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2017-04-27 FreeLogic/2017-Journal/article.tex
update

\title{Automating Free Logic in HOL, with an Application in Category Theory}
%\title{Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\title{Embedding Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\subtitle{(And an Exemplary Application in Category Theory)}

@@

\author{Christoph Benzm\"uller  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   University of Luxembourg, Luxembourg \& Freie Universit\"at Berlin, Berlin, Germany \\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of Califormia, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

  The purpose of this article is not to claim any novel results in
  category theory, but to demonstrate an elegant way to ``implement''
  and utilise interactive and automated reasoning in free logic, and
  to present respective experiments.

@@

\section{Introduction}
\label{intro}
Partiality and undefinedness are prominent challenges in various areas
of mathematics and computer science.  Unfortunately, however, modern
proof assistant systems and automated theorem provers based on
traditional classical or intuitionistic logics provide rather
inadequate support for these challenge concepts.  Free logic offers a
theoretically appealing solution, but is has been considered as rather
unsuited towards practical utilisation.

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free (or inclusive logic) in HOL. We
present an exemplary implementation of this idea in the mathematical
proof assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

To demonstrate the practical relevance of our new system, we present
in \S4 a stepwise development of axioms
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be implemented and validated within modern proof assistants such as
Isabelle/HOL.  We also address the relation of our axioms systems to
alternative proposals from the literature, including an axiom set
proposed by Freyd and Scedrov in their textbook ``Categories,
Allegories'' \cite{FreydScedrov90} for which we reveal a technical
flaw: either all operations, e.g. morphism composition, are total or
their axiom system is inconsistent. The repair for this problem is
quite straightforward, however. The solution essentially corresponds
to a set of axioms proposed by Scott \cite{Scott79} in the 1970s.

@@

Our exploration has been significantly supported by series of
experiments in which automated reasoning tools have been called from
within the proof assistant Isabelle/HOL via the
Sledgehammer tool. Moreover, we have obtained
very useful feedback at various stages from the model finder Nitpick
\cite{Nitpick} saving us from making several mistakes.

At the conceptual level this paper exemplifies a new style of
explorative mathematics which rests on a significant amount of
human-machine interaction with integrated interactive-auto\-ma\-ted
theorem proving technology. The experiments we have conducted are such
that the required reasoning is often too tedious and time-consuming
for humans to be carried out repeatedly with highest level of
precision. It is here where cycles of formalization and
experimentation efforts in Isabelle/HOL provided significant
support. Moreover, the technical inconsistency issue for axiom system
VII was discovered by automated theorem provers, which further
emphasises the added value of automated theorem proving in this area.

The content of article is combines, extends and clarifies the
contributions reported in two previous papers \cite{ICMS,ArXiv}.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic \cite{Lambert60,Scott67,lambert02:_free_logic} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subseteq \D$ is
chosen to characterize the subdomain of ``existing'' resp. ``defined''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

It is obvious how this can be used to model undefideness and
partiality: problematic terms, e.g. division by zero or improper
definite descriptions, still denote, but they refer to undefined
objects, that is, objects $d$ in $\D\setminus \ex$ lying outside of
the scope of quantification.
%In this framework partial and total functions are modelled as follows: 
Moreover, a function $f$ is \emph{total} if and only if for all $x$ we have $\ex x \imp \ex(f x)$ 
For \emph{partial} functions $f$ we may have some $x$ such that $\ex x$ but not
$\ex(f x)$. A function $f$ is \emph{strict}  if  and only if for all $x$
 we have $\ex(f x) \imp \ex x$.

The particular notion of free logic as exploited in the remainder of
this article has been proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.  
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.
\begin{figure}
\centering
\newcommand\firstellipse{(2,-5) ellipse (6cm and 4cm)}
\newcommand\secondellipse{(0,-5.3) ellipse (3.5cm and 2.5cm)}1
\begin{tikzpicture}
  \begin{scope}[fill opacity=1]
    \fill[gray!40] \firstellipse;
    \fill[gray!10] \secondellipse;
  \end{scope}
  \begin{scope}[very thick,font=\large]
    \draw \firstellipse node[font=\normalsize\bfseries] at (-.5,-5) {\underline{\textbf{$\ex$}: existing objects}};
    \draw \firstellipse node[font=\small] at (-.5,-5.5) {values of bound variables};
    \draw \secondellipse node[font=\normalsize\bfseries] at (4.3,-2.5) {\underline{\textbf{D}: raw objects}};
    \draw \secondellipse node[font=\small] at (4.3,-3) {values of free variables};
  \end{scope}
  \node[font=\normalsize\bfseries] at (6,-5) {$\star$};
  \node[font=\normalsize\bfseries] at (6,-5.3) {undefined};
\end{tikzpicture}
\caption{Illustration of the Semantical Domains of Free Logic \label{fig1}}
\end{figure}

We now formally introduce the syntax and semantics of Scott's \cite{Scott67} free logic as
assumed in the remainder of this article. We refer to this logic with $\FFOL$.

@@

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x\lambdot r$ are (compound) formulas of $\FFOL$
\item given a formula $r$ of $\FFOL$, then $\Descr x\lambdot r$ is a
  term of $\FFOL$ (definite  description)
\end{enumerate}
\end{definition}

Further formulas of $\FFOL$, including various defined notions of
equality, can be introduced as abbreviations.

  \emph{Substitution} of a term $r$ for a variable $x$
  in a term $s$ is denoted by $[r/x]s$.

@@

Regarding the semantics different options have been proposed in the literature. For
example, instead of a possible empty set of existing objects $\ex$, we
could postulate non-emptiness of $\ex$. Here we closely
follow the notion of free logic as proposed by Scott \cite{Scott67}.

\begin{definition}[Dual-domain semantics of $\FFOL$]
  A \emph{model structure for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g$, we
  define the interpretation function $\|\cdot\|^{M,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| q \|^{M,g} = I(q)$, where $q\in\P$ is an $0$-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{M,g} = \True$ iff $\| t_1 \|^{M,g} = \| t_2
  \|^{M,g}$ (this basic notion of primitive equality on $D$ implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in \S4, we will
  define and utilise further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}).
\item $\| p(t_1,\ldots,t_n) \|^{M,g} = \True$ iff $(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{M,g} = \True$ iff $\| r \|^{M,g} = \False$
\item $\| r \Impl s \|^{M,g} = \True$ iff $\| r \|^{M,g} = \False$
  or $\| s \|^{M,g} = \True$
\item $\| \forall x\lambdot r  \|^{M,g} = \True$ iff for all $d \in \ex$ we
  have $\|  r  \|^{M,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\begin{definition}[Validity]\label{ffollvalid}
  A formula $s_o$ is \emph{true} in model $M$ under assignment $g$ iff
  $\|s_o\|^{M,g} = T$; this is also denoted as $M,g \models^\FFOL s_o$.  A
  formula $s_o$ is called \emph{valid} in $M$, which is denoted as
  $M \models^\FFOL s_o$, iff $M,g \models^\FFOL s_o$ for all assignments
  $g$. Finally, a formula $s_o$ is called \emph{valid}, which we
  denote by $\models^\FFOL s_o$, iff $s_o$ is valid for all $M$. Moreover,
  we write $\Gamma\models^\FFOL \Delta$ (for sets of formulas $\Gamma$ and
  $\Delta$) iff there is a model $M$ and an assignment $g$ such that
  $M,g \models^\FFOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\FFOL t_o$ for
  at least one $t_o\in \Delta$.
\end{definition}

\subsection{Classical Higher-Order Logic}
Simple type theory, also referred to as classical higher-order logic
(HOL), is an expressive logic formalism that allows for higher-order
quantification, that is quantification over arbitrary set and function
variables.  It is based on the simply typed
$\lambda$-calculus\footnote{ For a detailed discussion of typed
  $\lambda$-calculi, we refer to the
  literature~\cite{DBLP:books/daglib/0032840}.  }
  %\footnote{
    %Although \sys~actually supports a polymorphic typed $\lambda$-calculus
    %we will
    %stick here to the simply typed $\lambda$-calculus for a better understanding.}
  and it has its origin in the work by
  Church~\cite{DBLP:journals/jsyml/Church40}.

\begin{definition}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{\mu}, \hol{\iota}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans,
  $\hol{\mu}$ is the type of individuals, and type
  $\hol{\iota}$ is employed as the type of possible worlds below. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The language of higher-order logic $\HOL$ with primitive equality and
choice is defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \cite{Andrews02}. The choice here is motivated by the
  signature supported in higher-order automated theorem provers such
  as LEO-II \cite{LEO-II} and Leo-III \cite{Leo-III}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ (for $\hol{\alpha\in T}$) and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (for $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{definition}

>>>>

\title{Automating Free Logic in HOL, with an Experimental Application in Category Theory}
%\title{Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\title{Embedding Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\subtitle{(And an Exemplary Application in Category Theory)}

@@

\author{Christoph Benzm\"uller  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   University of Luxembourg, Luxembourg \& Freie Universit\"at Berlin, Berlin, Germany \\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

  The purpose of this article is not to claim any novel results in
  category theory, but to demonstrate an elegant way to ``implement''
  and utilize interactive and automated reasoning in free logic, and
  to present respective experiments.

@@

\section{Introduction}
\label{intro}
Partiality and undefinedness are prominent challenges in various areas
of mathematics and computer science.  Unfortunately, however, modern
proof assistant systems and automated theorem provers based on
traditional classical or intuitionistic logics provide rather
inadequate support for these challenge concepts.  Free logic \cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} offers a
theoretically appealing solution, but is has been considered as rather
unsuited towards practical utilisation.

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free in HOL. We
present an exemplary implementation of this idea in the mathematical
proof assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

To demonstrate the practical relevance of our new system, we present
in \S4 a stepwise development of axioms
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be implemented and validated within modern proof assistants such as
Isabelle/HOL.  We also address the relation of our axiom systems to
alternative proposals from the literature, including an axiom set
proposed by Freyd and Scedrov in their textbook ``Categories,
Allegories'' \cite{FreydScedrov90} for which we reveal a technical
flaw: either all operations, e.g. morphism composition, are total or
their axiom system is inconsistent. The repair for this problem is
quite straightforward, however. The solution essentially corresponds
to a set of axioms proposed by Scott \cite{Scott79} in the 1970s.

@@

Our exploration has been significantly supported by series of
experiments in which automated reasoning tools have been called from
within the proof assistant Isabelle/HOL via the
Sledgehammer tool \cite{Sledgehammer}. Moreover, we have obtained
very useful feedback at various stages from the model finder Nitpick
\cite{Nitpick} saving us from making several mistakes.

At the conceptual level this paper exemplifies a new style of
explorative mathematics which rests on a significant amount of
human-machine interaction with integrated interactive-auto\-ma\-ted
theorem proving technology. The experiments we have conducted are such
that the required reasoning is often too tedious and time-consuming
for humans to be carried out repeatedly with highest level of
precision. It is here where cycles of formalization and
experimentation efforts in Isabelle/HOL provided significant
support. Moreover, the technical inconsistency issue for axiom system
VII was discovered by automated theorem provers, which further
emphasizes the added value of automated theorem proving in this area.

The content of article is based on the 
contributions reported in two previous papers \cite{ICMS,ArXiv}.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic (respectively
inclusive logic) 
\cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subseteq \D$ is
chosen to characterize the subdomain of ``\emph{existing}'' resp. ``\emph{defined}''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

It is obvious how this can be used to model undefideness and
partiality: problematic terms, e.g. division by zero or improper
definite descriptions, still denote, but they refer to undefined
objects, that is, objects $d$ in $\D\setminus \ex$ lying outside of
the scope of quantification.
%In this framework partial and total functions are modelled as follows: 
Moreover, a function $f$ is \emph{total} if and only if for all $x$ we
have $\ex x \imp \ex(f x)$.\footnote{The predication $\ex x$ represents that $x$ is a
  member of $E$.}
For \emph{partial} functions $f$ we may have some $x$ such that $\ex x$ but not
$\ex(f x)$. A function $f$ is \emph{strict}  if  and only if for all $x$
 we have $\ex(f x) \imp \ex x$.

The particular notion of free logic as exploited in the remainder of
this article has been proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.  
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.

\begin{figure}
\centering
\newcommand\firstellipse{(2,-5) ellipse (6cm and 4cm)}
\newcommand\secondellipse{(0,-5.3) ellipse (3.5cm and 2.5cm)}
\resizebox{.9\textwidth}{!}{ 
\begin{tikzpicture}
  \begin{scope}[fill opacity=1]
    \fill[gray!40] \firstellipse;
    \fill[gray!10] \secondellipse;
  \end{scope}
  \begin{scope}[very thick,font=\large]
    \draw \firstellipse node[font=\normalsize\bfseries] at (-.5,-5) {\underline{\textbf{$\ex$}: existing objects}};
    \draw \firstellipse node[font=\small] at (-.5,-5.5) {values of bound variables};
    \draw \secondellipse node[font=\normalsize\bfseries] at (4.3,-2.5) {\underline{\textbf{D}: raw objects}};
    \draw \secondellipse node[font=\small] at (4.3,-3) {values of free variables};
  \end{scope}
  \node[font=\normalsize\bfseries] at (6,-5) {$\star$};
  \node[font=\normalsize\bfseries] at (6,-5.3) {undefined};
\end{tikzpicture}
}
\caption{Illustration of the Semantical Domains of Free Logic \label{fig1}}
\end{figure}

We formally introduce the syntax and semantics of free logic as
assumed in the remainder of this article. We refer to this logic with $\FFOL$.

@@

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x\lambdot r$ are (compound) formulas of
  $\FFOL$, and
\item given a formula $r$ of $\FFOL$, then $\Descr x\lambdot r$ is a
  term of $\FFOL$ (definite  description).
\end{enumerate}
\end{definition}

Further terms and formulas of $\FFOL$, including various defined notions of
equality, can be introduced as abbreviations.

  % \emph{Substitution} of a term $r$ for a variable $x$
  % in a term $s$ is denoted by $[r/x]s$.

@@

Regarding the semantics different options have been proposed in the literature. For
example, instead of a possible empty set of existing objects $\ex$, we
could postulate non-emptiness of $\ex$. 
In fact, our approach below can be easily adapted for different variants of free and
inclusive logic.
Here we closely
follow the notion of free logic as proposed by Scott \cite{Scott67}.

\begin{definition}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. $\True$ or $\False$ denote truth and
  felshood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{definition}

\begin{definition}[Evaluation Function for $\FFOL$]
  The \emph{value} $\| s \|^{M,g}$ of a term or formula $s\in\FFOL$
  in a model $M=\langle D, \ex, I, \star\rangle$ under assignment $g$ 
  defined in the following way:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\setcounter{enumi}{\theenumTemp}
\item $\| q \|^{M,g} = I(q)$, where $q\in\P$ is an $0$-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{M,g} = \True$ if and only if $\| t_1 \|^{M,g} = \| t_2
  \|^{M,g}$ (this basic notion of primitive equality on $D$ implies
  that equations between ``undefined'' terms such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in \S4, we will
  define and utilize further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}, which behave differently).
\item $\| p(t_1,\ldots,t_n) \|^{M,g} = \True$ if and only if $(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{M,g} = \True$ if and only if $\| r \|^{M,g} = \False$
\item $\| r \Impl s \|^{M,g} = \True$ if and only if $\| r \|^{M,g} = \False$
  or $\| s \|^{M,g} = \True$
\item $\| \forall x\lambdot r  \|^{M,g} = \True$ if and only if for all $d \in \ex$ we
  have $\|  r  \|^{M,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\begin{definition}[Validity]\label{ffollvalid}
  A formula $s_o$ is \emph{true} in model $M$ under assignment $g$ if and only if
  $\|s_o\|^{M,g} = T$; this is also denoted as $M,g \models^\FFOL s_o$.  A
  formula $s_o$ is called \emph{valid} in $M$, which is denoted as
  $M \models^\FFOL s_o$, if and only if $M,g \models^\FFOL s_o$ for all assignments
  $g$. Finally, a formula $s_o$ is called \emph{valid}, which we
  denote by $\models^\FFOL s_o$, if and only if $s_o$ is valid for all
  $M$. 
% Moreover,
%   we write $\Gamma\models^\FFOL \Delta$ (for sets of formulas $\Gamma$ and
%   $\Delta$) if and only if there is a model $M$ and an assignment $g$ such that
%   $M,g \models^\FFOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\FFOL t_o$ for
%   at least one $t_o\in \Delta$.
\end{definition}

\subsection{Classical Higher-Order Logic}
Simple type theory, also referred to as classical higher-order logic
(HOL) \cite{AndrewsSEP}, is an expressive logic formalism which is based on the simply typed
$\lambda$-calculus~\cite{barendregt13book}. 
  %\footnote{
    %Although \sys~actually supports a polymorphic typed $\lambda$-calculus
    %we will
    %stick here to the simply typed $\lambda$-calculus for a better understanding.}
  HOL has its origin in the work by
  Church~\cite{Church40}.

For a detailed discussion of the syntax, semantics and automation
of  HOL we refer to the literature (see e.g.
\cite{AndrewsSEP,B5,J6} and the
references therein). Below we introduce a variant of \HOL with primitive equality
and description.

\begin{definition}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{i}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans and 
  $\hol{i}$ is the type of individuals. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The language of higher-order logic $\HOL$ with primitive equality and
description is defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III} and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{definition}

We assume familiarity with $\lambda$-conversion (e.g. \emph{$\alpha$-renaming} and
\emph{$\beta\eta$-reduction}) and \emph{$\beta\eta$-normal forms}~\cite{barendregt13book}.
